---
title: "From Simple to Multiple Regression in Global Health"
subtitle: "Confounding, Causality, and AI Perspectives (One Health)"
author: "Pinto Jonathan Katende"
email: "jonathankatende.pinto@ubep.unipd.it"
date: last-modified
format:
  html:
    toc: true
    toc-float: true
    code-fold: true
    code-summary: "View R Code"
    number-sections: false
execute:
  warning: false
  message: false
---

## 1. Introduction

### 1.1 Research Question

> Does increasing environmental exposure (e.g., air pollution) increase disease risk?

We begin with a simple model, then progressively uncover hidden biases and complexity.

---

## 2. Data Simulation: A One-Health Perspective

```{r}
#| label: setup
library(knitr)
library(kableExtra)
library(tidyverse)
library(gtsummary)
library(car)
library(lmtest)
library(ggeffects)
library(ggpubr)
library(performance)
library(see)
library(corrplot)

generate_complex_data <- function(n = 200, seed = 42){
  set.seed(seed)
  
  age <- round(runif(n, 20, 80))
  BMI <- round(rnorm(n, 25, 4), 1)
  comorbidity <- sample(c("High", "Low"), n, replace = TRUE)
  comorb_numeric <- ifelse(comorbidity == "High", 1, 0)
  
  # One Health exposure (PM2.5 air pollution)
  air_pollution_level <- round(5 + (0.15 * age) + (0.4 * BMI) + 
                         (4 * comorb_numeric) + rnorm(n, 0, 3), 1)
  
  # Health outcome
  disease_risk <- round(10 + (0.5 * air_pollution_level) +
                  (-0.2 * age) +
                  (-5 * comorb_numeric) +
                  rnorm(n, 0, 2), 1)
  
  data.frame(age, BMI, comorbidity, air_pollution_level, disease_risk)
}

data <- generate_complex_data()

data$comorbidity <- ifelse(data$comorbidity == "High", "Low", "High")
data$comorbidity <- factor(data$comorbidity, levels = c("Low", "High"))
  
kable(head(data), caption = "Table 1: Baseline characteristics of the simulated cohort") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```

---

## 3. The Naive Approach

### 3.1 Simple Linear Regression

```{r}
#| label: simple-model
model_simple <- lm(disease_risk ~ air_pollution_level, data = data)
summary(model_simple)
```

---

### 3.2 Visualization

```{r}
#| label: plot-simple
ggplot(data, aes(air_pollution_level, disease_risk)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Naive Relationship: Air Pollution vs Disease Risk") +
  theme_minimal()
```

---

## 4. Hidden Structure: Confounding

### 4.1 Correlation Between Predictors

```{r}
#| label: correlation
#| fig-width: 8
#| fig-height: 8

cor(data$air_pollution_level, data$age)
cor(data$age, data$disease_risk)


data_num <- data %>% select(-comorbidity)
cor_matrix <- cor(data_num)
testRes <- psych::corr.test(data_num, adjust = "none")

corrplot(cor_matrix,
         p.mat = testRes$p, 
         method = 'circle', 
         type = 'lower', 
         insig = 'blank', 
         addCoef.col = 'black', 
         number.cex = 0.8, 
         order = 'AOE', 
         diag = FALSE)
```

> Age is moderately correlated with air pollution exposure (r = 0.54) and strongly associated with disease risk (r = -0.62), indicating a substantial confounding effect. Adjusting for age is therefore essential to obtain an unbiased estimate of the exposure–outcome relationship

---

### 4.2 Interpretation

Age influences both:

* exposure (air pollution)
* outcome (disease risk)

  This creates **confounding bias**

---

## 5. The Adjustment

### 5.1 Multiple Linear Regression

```{r}
#| label: multiple-model
model_full <- lm(disease_risk ~ air_pollution_level + age + BMI + comorbidity, data = data)

tbl_merge(
  list(
    tbl_regression(model_simple),
    tbl_regression(model_full)
  ),
  tab_spanner = c("**Naive Model**", "**Adjusted Model**")
)
```

---

### 5.2 Key Insight

If the coefficient of air pollution changes → **confounding exists**

---

## 6. Model Refinement

### 6.1 Variable Selection (AIC)

```{r}
#| label: selection

model_full <- lm(disease_risk ~ air_pollution_level + age + BMI + comorbidity, data = data)

# Backward selection
model_final <- step(model_full, direction = "backward")

summary(model_final)
```

---

### AI Perspective

> Stepwise selection is a **primitive machine learning approach**.

Modern AI methods use:

* **Lasso (L1 regularization)**
* **Ridge (L2 regularization)**

These methods:

* handle high-dimensional data
* reduce overfitting
* perform automatic feature selection

> While step() chooses which variables to keep, the Lasso "overwrites" unnecessary variables to zero, which is much more efficient when you have 200 predictors instead of 4.

---

## 7. Model Diagnostics

### 7.1 Global Diagnostic

```{r}
#| label: diagnostics
#| fig-width: 10
#| fig-height: 12
check_model(model_final)
```

  This replaces multiple manual plots with a **clean, unified diagnostic panel**

---

### 7.2 Statistical Tests

```{r}
#| label: tests

#independence of residuals
dwtest(model_final)

# Homoscedasticity
bptest(model_final)

# Normality
shapiro.test(resid(model_final))

# Multicollinearity
vif(model_final)
```

---

## 8. Interaction Effects

### 8.1 Motivation

Does pollution affect everyone equally?

  Maybe not - it may depend on age.

---

### 8.2 Simulation with Interaction

```{r}
#| label: interaction-data

generate_interaction_data <- function(n = 250, seed = 42){
  set.seed(seed)
  
  age <- round(runif(n, 20, 80))
  comorbidity <- sample(c("Low", "High"), n, replace = TRUE)
  comorb_numeric <- ifelse(comorbidity == "High", 1, 0)
  
  air_pollution_level <- 5 + (0.3 * age) + (4 * comorb_numeric) + rnorm(n, 0, 3)
  
  disease_risk <- 10 + (0.7 * air_pollution_level) +
                  (-0.1 * age) +
                  (-0.01 * air_pollution_level * age) +
                  (-5 * comorb_numeric) +
                  rnorm(n, 0, 2)
  
  data.frame(age, comorbidity, air_pollution_level, disease_risk)
}

data_inter <- generate_interaction_data()

data_inter$comorbidity <- ifelse(data_inter$comorbidity == "High", "Low", "High")
data_inter$comorbidity <- factor(data_inter$comorbidity, levels = c("Low", "High"))
```

---

### 8.3 Interaction Model

```{r}
#| label: interaction-model

model_inter <- lm(disease_risk ~ air_pollution_level * age + comorbidity, data = data_inter)

model_inter %>% 
  tbl_regression(
    label = list(
      air_pollution_level ~ "Air Pollution",
      age ~ "Age",
      `air_pollution_level:age` ~ "Pollution × Age Interaction"
    )
  ) %>%
  bold_p()
```

---

### 8.4 Visualization

```{r}
#| label: interaction-plot

data_inter %>%
  mutate(age_group = cut(age, breaks = 3)) %>%
  ggplot(aes(x = air_pollution_level, y = disease_risk, color = age_group)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal()
```

---

## 9. Marginal Effects (AI-friendly interpretation)

```{r}
#| label: ggeffects

pred <- ggpredict(model_inter, terms = c("air_pollution_level", "age [30,50,70]"))

plot(pred) +
  labs(
    title = "Marginal Effect of Air Pollution by Age",
    subtitle = "Effect decreases with increasing age",
    x = "Air Pollution (PM2.5)",
    y = "Predicted Disease Risk"
  )
```

---

## 10. Conclusion

We have shown that:

* Simple models can be misleading
* Confounding biases estimates
* Multiple regression corrects bias
* Interaction reveals population heterogeneity
* AI methods extend these ideas to complex datasets

---

## Final Takeaway

> A good model is not the simplest one,
> but the one that reflects the true data-generating process.
