---
title: "Logistic Regression: Diabetes Prediction"
subtitle: "From Full Models to Optimized Signal"
author: "Jonathan Pinto"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show Code"
    theme: cosmo
    highlight-style: dracula
execute:
  echo: true
  warning: false
  message: false
---

## 1. Introduction {background-color="#2c3e50"}

In Logistic Regression, we model the probability of a binary outcome (0 or 1).

- **Outcome**: Diabetes (1) or No Diabetes (0).
- **Goal**: Identify which clinical factors (Glucose, BMI, Age, etc.) drive the risk.
- **Method**: Likelihood ratio tests and AIC for selection.

## 2. Setup & Data Exploration

```{r}
#| label: setup
library(tidyverse)
library(gtsummary)
library(caret)
library(broom)
library(pROC)
library(dplyr)
library(ggplot2)

data <- read.csv("diabetes.csv")

# Ensure Outcome is a factor
data <- data %>%
  mutate(Outcome = factor(Outcome, levels = c(0, 1), labels = c("No", "Yes")))
```

## 3. Clinical Table 1

```{r}
#| label: table1
data %>%
  tbl_summary(by = Outcome) %>%
  add_p() %>%
  bold_labels()
```

## 4. The "Noisy" Full Model

```{r}
#| label: full-model

full_model_all <- glm(Outcome ~ ., data = data, family = binomial)

full_model_all %>% 
  tbl_regression(exponentiate = TRUE) %>%
  bold_p()
```

## 5. Model Optimization: Forward Selection

```{r}
#| label: forward-selection
#| results: hide

# Null model
null_model <- glm(Outcome ~ 1, data = data, family = binomial)

# Full model (defined explicitly)
full_model_defined <- glm(
  Outcome ~ Pregnancies + Glucose + BloodPressure + SkinThickness + 
    Insulin + BMI + DiabetesPedigreeFunction + Age, 
  data = data, 
  family = binomial
)

# Forward selection
forward_model <- step(
  null_model, 
  scope = list(lower = null_model, upper = full_model_defined), 
  direction = "forward",
  trace = 0
)
```

## Comparison: Full vs Forward Model

```{r}
#| label: comparison

full_table <- full_model_defined %>% 
  tbl_regression(exponentiate = TRUE)

forward_table <- forward_model %>% 
  tbl_regression(exponentiate = TRUE) %>% 
  bold_labels()

tbl_merge(
  list(full_table, forward_table),
  tab_spanner = c("**Full Model (Initial)**", "**Forward Selected Model**")
)
```

## Visualizing Odds Ratios

```{r}
#| label: or-plot
#| fig-cap: "Forest Plot of Odds Ratios (Forward Model)"

plot_data <- tidy(forward_model, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)")

ggplot(plot_data, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +
  geom_point(size = 3, color = "#2E9FDF") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  scale_x_log10() +
  labs(
    title = "Odds Ratios: Forward Selection Results",
    subtitle = "Variables selected based on AIC optimization",
    x = "Odds Ratio (Log Scale)",
    y = ""
  ) +
  theme_minimal()
```

## Results summary

```{r}
#| label: forward-summary

summary(forward_model)
```

## 6. Model Validation (ROC Curve)

```{r}
#| label: roc-curve

# Predicted probabilities
probs <- predict(forward_model, type = "response")

# ROC
roc_obj <- roc(data$Outcome, probs)

plot(
  roc_obj, 
  col = "#2E9FDF", 
  lwd = 3,
  main = paste("ROC Curve (AUC =", round(auc(roc_obj), 3), ")")
)
```

## 7. Calibration

```{r}
#| label: calibration

# 1. Use the correct model name (forward_model)
data$pred <- predict(forward_model, type = "response")

# 2. Convert the Factor "Outcome" to numeric 0/1 for calculating means
calib_data <- data %>%
  mutate(
    # In R, factors No/Yes are stored as 1/2. Subtracting 1 gives 0/1.
    outcome_num = as.numeric(Outcome) - 1,
    bin = ntile(pred, 10)
  ) %>%
  group_by(bin) %>%
  summarise(
    pred_mean = mean(pred),
    obs_mean = mean(outcome_num)
  )

# 3. Plot
ggplot(calib_data, aes(x = pred_mean, y = obs_mean)) +
  geom_point(size = 3, color = "#2E9FDF") +
  geom_line(color = "#2E9FDF") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Calibration Plot",
    subtitle = "Predicted vs. Observed Diabetes Prevalence",
    x = "Predicted Probability",
    y = "Observed Probability"
  ) +
  theme_minimal()
```

