---
title: "From Linear Models to GAM"
subtitle: "Understanding Non-Linearity in Global Health and One Health Systems"
author: "Jonathan Katende Pinto"
date: last-modified
format:
  pdf:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
execute:
  warning: false
  message: false
---

## 1. Introduction: The Illusion of Linearity

We often start with a simple assumption:

> Increasing environmental exposure leads to proportional increase in health risk.

This naturally leads to linear regression.

However, environmental and epidemiological systems are rarely linear.

---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(mgcv)
library(gtsummary)
library(gratia)
library(knitr)
library(performance)
```

## 2. Simulating an Environmental Health Reality

```{r}
#| label: simulation-visual

set.seed(123)

n <- 150

air_pollution_level <- runif(n, 0, 100)

max_population_risk <- 100
exposure_half_risk <- 20

respiratory_risk <- 
  (max_population_risk * air_pollution_level) /
  (exposure_half_risk + air_pollution_level) +
  rnorm(n, 0, 5)

df <- data.frame(
  air_pollution_level,
  respiratory_risk
)

ggplot(df, aes(air_pollution_level, respiratory_risk)) +
  geom_point(alpha = 0.5) +

  geom_smooth(
    aes(color = "True pattern (loess)"),
    method = "loess",
    se = FALSE,
    size = 1.2
  ) +

  geom_smooth(
    aes(color = "Linear model"),
    method = "lm",
    formula = y ~ x,
    linetype = "dashed",
    se = FALSE
  ) +

  scale_color_manual(
    name = "Model",
    values = c(
      "True pattern (loess)" = "blue",
      "Linear model" = "red"
    )
  ) +

  labs(
    title = "Environmental Exposure and Health Risk",
    x = "Air Pollution Level",
    y = "Respiratory Risk"
  ) +

  theme_minimal()
```

---

## 3. The Wrong Model: Linear Regression

The linear model assumes:

$$
Y = \beta_0 + \beta_1 X + \varepsilon
$$

This implies a constant increase in health risk for each additional unit of environmental exposure.

```{r}
#| label: lm

model_lm_simple <- lm(
  respiratory_risk ~ air_pollution_level,
  data = df
)

ggplot(df, aes(air_pollution_level, respiratory_risk)) +
  geom_point() +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Linear Model: Incorrect Assumption",
    x = "Air Pollution Level",
    y = "Respiratory Risk"
  ) +
  theme_minimal()

summary(model_lm_simple)
```

---

## 4. The Diagnostic: Residuals

```{r}
#| label: residuals

ggplot(model_lm_simple, aes(.fitted, .resid)) +
  geom_point() +
  geom_smooth(method = "loess", color = "red") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Residual Pattern",
    subtitle = "Non-random pattern indicates model misspecification",
    x = "Fitted values",
    y = "Residuals"
  ) +
  theme_minimal()
```

> Residuals are not just diagnostics — they often reveal when the model is misspecified.

A correct model should produce residuals randomly scattered around zero.

---

## 5. Polynomial Regression

To capture curvature, we extend the linear model:

$$
Y = \beta_0 + \beta_1 X + \beta_2 X^2 + \varepsilon
$$

This allows the model to bend, but it still does not represent the true environmental mechanism.

```{r}
#| label: polynomial

model_poly <- lm(
  respiratory_risk ~ air_pollution_level + I(air_pollution_level^2),
  data = df
)

ggplot(df, aes(air_pollution_level, respiratory_risk)) +
  geom_point() +
  geom_smooth(
    method = "lm",
    formula = y ~ x + I(x^2),
    color = "purple"
  ) +
  labs(
    title = "Polynomial Regression",
    x = "Air Pollution Level",
    y = "Respiratory Risk"
  ) +
  theme_minimal()

summary(model_poly)
```

Polynomial regression improves the fit by introducing curvature, but it remains a statistical approximation.

Although the model captures non-linearity, it introduces an artificial decline in risk at high exposure levels, which is not epidemiologically plausible.

> A good fit does not imply a correct model.

---

### Interpretation

The fitted model can be written as:

$$
Risk = 19.33 + 1.51(Exposure) - 0.009(Exposure^2)
$$

* The positive linear term indicates an initial increase in health risk.
* The negative quadratic term introduces curvature and diminishing returns.

The model predicts a maximum risk at:

$$
Exposure_{max} = \frac{-\beta_1}{2\beta_2}
$$

However, the decline beyond this point is artificial and does not reflect the true exposure–risk mechanism.

---

## 6. Model Comparison

### 6.1 ANOVA Comparison

```{r}
#| label: anova-lm

model_base <- lm(
  respiratory_risk ~ air_pollution_level,
  data = df
)

model_extended <- lm(
  respiratory_risk ~ air_pollution_level + I(air_pollution_level^2),
  data = df
)

anova(model_base, model_extended)
```

### Interpretation

* p-value < 0.05 → the additional term improves the model
* p-value > 0.05 → the simpler model is sufficient

---

## 7. The Temptation: Transformations

```{r}
#| label: log-transform

df_log <- df %>%
  mutate(log_exposure = log(air_pollution_level + 1))

model_log <- lm(
  respiratory_risk ~ log_exposure,
  data = df_log
)

ggplot(df_log, aes(log_exposure, respiratory_risk)) +
  geom_point() +
  geom_smooth(method = "lm", color = "purple") +
  labs(
    title = "Log Transformation",
    x = "Log Exposure",
    y = "Respiratory Risk"
  ) +
  theme_minimal()
```

### Why transformations can fail

* They distort interpretation
* They may introduce heteroscedasticity
* They force the data into a linear structure

---

## 8. Non-linear Least Squares (NLS)

Instead of transforming the data, we directly model the environmental mechanism:

$$
Risk = \frac{R_{max} \cdot Exposure}{E_{50} + Exposure} + \varepsilon
$$

Where:

* (R_{max}): maximum achievable population risk
* (E_{50}): exposure level associated with 50% of the maximum risk

```{r}
#| label: nls

model_nls <- nls(
  respiratory_risk ~ 
    (Rmax * air_pollution_level) /
    (E50 + air_pollution_level),
  data = df,
  start = list(Rmax = 80, E50 = 10)
)

summary(model_nls)
```

```{r}
#| label: nls-plot

df$pred_nls <- predict(model_nls)

ggplot(df, aes(air_pollution_level, respiratory_risk)) +
  geom_point() +
  geom_line(aes(y = pred_nls), color = "green", size = 1) +
  labs(
    title = "NLS Model: Correct Environmental Fit",
    x = "Air Pollution Level",
    y = "Respiratory Risk"
  ) +
  theme_minimal()
```

---

## 9. Other Non-linear Forms

### Exponential Model

$$
Y = a e^{bX}
$$

Used for epidemic growth or infectious disease propagation.

```{r}
#| label: exp

model_exp <- nls(
  respiratory_risk ~ a * exp(b * air_pollution_level),
  data = df,
  start = list(a = 1, b = 0.01)
)

summary(model_exp)
```

---

### Power Model

$$
Y = a X^b
$$

Common in environmental scaling relationships.

```{r}
#| label: power

model_power <- nls(
  respiratory_risk ~ a * air_pollution_level^b,
  data = df,
  start = list(a = 1, b = 1)
)

summary(model_power)
```

---

### Logarithmic Model

$$
Y = a \log(X)
$$

Represents rapid initial increase followed by slowing growth.

```{r}
#| label: log-model

model_log2 <- nls(
  respiratory_risk ~ a * log(air_pollution_level + 1),
  data = df,
  start = list(a = 10)
)

summary(model_log2)
```

---

## 9.1 Self-Starting Non-linear Models

In practice, choosing starting values manually can be difficult.

To simplify this process, R provides several self-starting non-linear models (`SS*`) that automatically estimate reasonable initial values from the data.

This reduces convergence problems and simplifies model fitting.

```{r}
#| label: nls-auto

model_auto <- nls(
  respiratory_risk ~ 
    SSmicmen(air_pollution_level, Rmax, E50),
  data = df
)

summary(model_auto)
```

The function `SSmicmen()` implements the Michaelis–Menten model:

$$
Y = \frac{R_{max} \cdot X}{E_{50} + X}
$$

but automatically estimates starting values before optimization begins.

This is particularly useful in epidemiology and environmental health studies where plausible parameter values may not be known in advance.


>Self-starting models improve convenience, but they do not eliminate all convergence problems.

> If the data are highly noisy, sparse, or poorly scaled, optimization may still fail.

---


## 10. Model Comparison

```{r}
#| label: aic

data.frame(
  Model = c(
    "Linear",
    "Polynomial",
    "Log",
    "NLS",
    "Exponential",
    "auto-NLS"
  ),
  AIC = c(
    AIC(model_lm_simple),
    AIC(model_poly),
    AIC(model_log),
    AIC(model_nls),
    AIC(model_exp),
    AIC(model_auto)
  )
)
```

### Interpretation

* Lower AIC indicates a better balance between fit and complexity.
* Large AIC differences suggest structural misspecification.

---

## 11. GAM: Flexible Non-linear Modeling

What if we do not know the true functional form?

Instead of imposing a structure, GAM learns it directly from the data.

---

### 11.1 Complex Environmental Signal

```{r}
#| label: gam-simulation

set.seed(123)

n <- 400

air_pollution_level <- runif(n, -3, 3)

noise <- rnorm(n, 0, 1)

respiratory_risk <- 
  4 * sin(3 * air_pollution_level) +
  cos(0.8 * air_pollution_level) -
  1.5 * (air_pollution_level^2) +
  noise

df_gam <- data.frame(
  air_pollution_level,
  respiratory_risk
)
```

This combines:

* seasonal environmental fluctuations
* epidemic waves
* long-term population trends

---

### 11.2 Linear Regression Failure

```{r}
#| label: gam-lm

model_lm_complex <- lm(
  respiratory_risk ~ air_pollution_level,
  data = df_gam
)

ggplot(df_gam, aes(air_pollution_level, respiratory_risk)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", color = "red", size = 1.2) +
  labs(
    title = "Linear Regression: Total Failure",
    subtitle = "The model misses oscillations and curvature"
  ) +
  theme_minimal()
```

---

### 11.3 The GAM Idea

$$
Y = \beta_0 + s(X) + \varepsilon
$$

Where:

* (s(X)) is a smooth function estimated from the data.

---

### 11.4 GAM Fit

```{r}
#| label: gam-model

model_gam_complex <- gam(
  respiratory_risk ~ s(air_pollution_level),
  data = df_gam,
  method = "REML"
)

df_gam$pred_gam <- predict(model_gam_complex)
```

```{r}
#| label: gam-plot

ggplot(df_gam, aes(air_pollution_level, respiratory_risk)) +
  geom_point(alpha = 0.15) +
  geom_line(aes(y = pred_gam), color = "cyan4", size = 1.5) +
  geom_smooth(
    method = "lm",
    color = "red",
    linetype = "dashed",
    se = FALSE
  ) +
  labs(
    title = "GAM Recovery",
    subtitle = "The spline captures hidden environmental structure"
  ) +
  theme_minimal()
```

---

### 11.5 Smooth Function Visualization

```{r}
#| label: gam-diagnostics

draw(model_gam_complex) +
  labs(
    title = "Estimated Smooth Function",
    subtitle = "Recovered hidden structure"
  )
```

---

### 11.6 Model Comparison

```{r}
#| label: gam-comparison

results <- data.frame(
  Model = c("Linear Model", "GAM"),
  Adjusted_R2 = c(
    summary(model_lm_complex)$adj.r.squared,
    summary(model_gam_complex)$r.sq
  ),
  AIC = c(
    AIC(model_lm_complex),
    AIC(model_gam_complex)
  )
)

kable(results, digits = 3)
```

---

### Interpretation

* Linear model → global trend only
* GAM → local + global structure

> GAM adapts to the data instead of forcing a predefined shape.

---

## 12. GAM with Multiple Predictors

A hybrid GAM combines smooth and linear terms:

$$
Y = \beta_0 + s(X_1) + \beta_2 X_2 + \beta_3 X_3 + \varepsilon
$$

---

### 12.1 Simulation

```{r}
#| label: gam-multi

set.seed(42)

n <- 200

air_pollution_level <- runif(n, 0, 100)

age <- runif(n, 20, 80)

temperature <- rnorm(n, 25, 5)

respiratory_risk <-
  4 * sin(air_pollution_level / 10) +
  cos(air_pollution_level / 20) -
  0.02 * air_pollution_level^2 +
  (-0.2 * age) +
  rnorm(n, 0, 2)

df2 <- data.frame(
  air_pollution_level,
  age,
  temperature,
  respiratory_risk
)
```

---

### 12.2 Wrong Approach: Smoothing Everything

```{r}
#| label: gam-all

gam_all <- gam(
  respiratory_risk ~ 
    s(air_pollution_level) +
    s(age) +
    s(temperature),
  data = df2
)

plot(gam_all, pages = 1)
```

### Problem

* unnecessary complexity
* reduced interpretability
* not all variables require smoothing

---

### 12.3 Hybrid GAM

```{r}
#| label: gam-hybrid

gam_hybrid <- gam(
  respiratory_risk ~
    s(air_pollution_level) +
    age +
    temperature,
  data = df2
)

plot(gam_hybrid, shade = TRUE)
```

---

### 12.4 Interpretation

```{r}
#| label: gam-summary

summary(gam_hybrid)
```

Interpretation of effective degrees of freedom (edf):

* edf ≈ 1 → approximately linear
* edf > 1 → non-linear relationship


> Because the data were simulated from a highly structured non-linear process, the GAM is able to recover almost the entire signal. In real epidemiological datasets, performance is usually lower due to measurement error, unmeasured confounding, and population heterogeneity.

---

### 12.5 Diagnostics

```{r}
#| label: gam-check
#| fig-width: 10
#| fig-height: 8

appraise(model_gam_complex)

gam.check(model_gam_complex)

appraise(gam_hybrid)

gam.check(gam_hybrid)
```

These diagnostics evaluate:

* residual structure
* adequacy of smoothing
* model assumptions

---

## 12.6 Tensor Product Smooths: Joint Non-linear Effects

```{r}
#| label: tensor-simulation

set.seed(123)

n <- 300

air_pollution_level <- runif(n, 0, 100)

temperature <- runif(n, 10, 40)

respiratory_risk <-
  0.03 * air_pollution_level^1.2 +
  0.04 * temperature^2 +
  0.02 * air_pollution_level * temperature +
  rnorm(n, 0, 3)

df_tensor <- data.frame(
  air_pollution_level,
  temperature,
  respiratory_risk
)
```

### Tensoriel GAM
```{r}
#| label: tensor-gam

model_tensor <- gam(
  respiratory_risk ~ 
    te(air_pollution_level, temperature),
  data = df_tensor,
  method = "REML"
)

summary(model_tensor)
```


```{r}
#| label: tensor-plot

vis.gam(
  model_tensor,
  type = "response",
  plot.type = "persp",
  phi = 30,
  theta = 30,
  n.grid = 50,
  color = "terrain"
)
```

The surface plot reveals that respiratory risk is not determined by pollution or temperature independently.

Instead, the risk emerges from their interaction:

high pollution combined with high temperature produces the highest risk,
while moderate exposure levels may have limited impact.

This illustrates a core principle of modern epidemiology:

health outcomes often emerge from interacting systems rather than isolated variables.

---

## 12.7 Common Diagnostic Problems in GAMs

Model diagnostics are not merely technical checks.

They help identify:
- structural misspecification,
- distributional problems,
- overfitting,
- insufficient smoothing,
- and violations of statistical assumptions.

| Diagnostic Issue | Visual Symptom | Statistical Interpretation | Corrective Action |
|---|---|---|---|
| Non-linearity | Patterns or “waves” in residuals vs fitted plot | The model misses structural complexity (underfitting) | Increase basis dimension (`k`) or add non-linear terms |
| Heteroscedasticity | “Fan” or “funnel” shape in residuals | Non-constant variance | Use another distribution family (`Gamma`, `nb`) or transform the response |
| Non-normality | QQ-plot strongly deviates from the line | Residual distribution is skewed or heavy-tailed | Investigate outliers or consider robust distributions |
| Autocorrelation | Residual clustering across time or space | Violation of independence assumption | Use GAMM with correlation structures or spatial effects |
| Under-smoothing | Low p-value in `gam.check()` (`k-index < 1`) | Basis dimension too small | Increase `k` |
| Overfitting | Very high edf with poor generalization | Model captures noise instead of signal | Increase smoothing penalty (`gamma`) or reduce `k` |

---

## 12.8 Understanding Basis Functions in GAMs

In GAMs, smooth functions are constructed using basis functions (`bs`).

The default basis in `mgcv` is:

```r
bs = "tp"
```

which stands for **thin plate regression splines**.

Thin plate splines are widely used because they:

* do not require manual knot placement,
* are generally stable and flexible,
* work well as general-purpose smoothers,
* automatically balance flexibility and smoothness.

### Common Alternative Bases

| Basis  | Meaning                     | Main Idea                                                                   |
| ------ | --------------------------- | --------------------------------------------------------------------------- |
| `"tp"` | Thin plate spline           | Default general-purpose smoother                                            |
| `"cr"` | Cubic regression spline     | Faster computation with fixed knots                                         |
| `"cs"` | Cubic spline with shrinkage | Allows unnecessary smooth terms to shrink toward zero                       |
| `"ps"` | Penalized spline            | Flexible smoothing with explicit penalty control                            |
| `"ds"` | Duchon spline               | Generalization of thin plate splines for complex multidimensional smoothing |

---

## Interaction Smooths in GAMs

Instead of a simple smooth term:

```r
s(x)
```

GAMs also allow smooth interactions between variables.

### Common Interaction Terms

| Function   | Interpretation                                                      |
| ---------- | ------------------------------------------------------------------- |
| `te(x, z)` | Main smooth effects + interaction between `x` and `z`               |
| `ti(x, z)` | Interaction smooth only                                             |
| `t2(x, z)` | Advanced tensor-product smooth with more flexible penalty structure |

### Conceptual Difference

* `te()` models both the individual effects and their interaction.
* `ti()` isolates only the interaction component.
* `t2()` is mainly used in advanced GAM formulations requiring complex penalty decomposition.

---

## 13. Final Takeaway

* Linear models are often insufficient in environmental and epidemiological systems.
* Transformations may distort interpretation.
* NLS models respect mechanistic relationships.
* GAM provides flexibility when the functional form is unknown.
* Hybrid GAMs balance interpretability and flexibility.

---

### Final Message

> The goal is not to fit the simplest model,
> but the model that best represents the underlying environmental and epidemiological mechanism.

> In Global Health and One Health research, non-linearity is the rule rather than the exception.
