---
title: "Practical Survival Analysis: Heart Failure Study"
subtitle: "A Biostatistical Approach for Global Health, AI and One Health (GIAHE)"
author: "Jonathan Pinto"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
execute:
  warning: false
  message: false
---

## 1. Introduction

In Global Health and the **One Health framework**, survival analysis is essential to study:

- time to death  
- time to disease emergence  
- duration of epidemics
- time to events

---

## 2. Setup and Data

```{r}
#| label: setup

library(survival)
library(survminer)
library(tidyverse)
library(gtsummary)
library(rms)
library(knitr)
library(kableExtra)

heart <- read.csv("heart_failure_clinical_records_dataset.csv")

heart <- heart %>%
  mutate(
    DEATH_EVENT = as.numeric(DEATH_EVENT),
    status_label = ifelse(DEATH_EVENT == 1, "Death", "Censored"),
    age_group = ifelse(age >= 65, "Senior (>=65)", "Junior (<65)"),
    low_ef = ifelse(ejection_fraction < 40, "Low EF (<40%)", "Normal EF")
  )


heart %>%
  head() %>%
  kable(caption = "Preview of the Heart Failure Clinical Dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = F, 
                position = "left") %>%
  scroll_box(width = "100%")

surv_obj <- Surv(time = heart$time, event = heart$DEATH_EVENT)
```

---

## 3. Kaplan-Meier: Overall Survival

```{r}
#| label: km-overall

km_overall <- survfit(surv_obj ~ 1, data = heart)

print(km_overall)

ggsurvplot(
  km_overall,
  data = heart,
  conf.int = TRUE,
  risk.table = TRUE,
  palette = "black",
  title = "Overall Survival Curve",
  xlab = "Time (Days)",
  ylab = "Survival Probability"
)
```

```{r}
#| label: km-100days

summary_100 <- summary(km_overall, times = 100)

data.frame(
  time = summary_100$time,
  survival = summary_100$surv,
  lower_ci = summary_100$lower,
  upper_ci = summary_100$upper
)

ggsurvplot(
  km_overall,
  data = heart,
  conf.int = TRUE,
  risk.table = TRUE,
  palette = "black",
  title = "Overall Survival Curve",
  xlab = "Time (Days)",
  ylab = "Survival Probability"
)$plot +
  geom_vline(xintercept = 100, linetype = "dashed", color = "red") +
  annotate("text", x = 100, y = 0.2, label = "100 days", color = "red", angle = 90)
```

**The "NA" Median Survival Mystery**

When reviewing the summary of the overall survival model, you will notice that the output for the median survival time returns NA. This is a frequent occurrence in epidemiological studies and is important to understand:

- **Statistical Definition**: The Median Survival Time is the exact time point when the survival probability ($S(t)$) drops to 0.5 (50%).

- **The Data Reality**: In this heart failure cohort, the survival probability at the end of the follow-up period (Day 241) is approximately 0.576.

- **The Conclusion**: Because the curve never reaches the 50% threshold, the median is "not reached". This indicates that more than half of the study population was still alive or censored at the time the study concluded.

**The Solution: Landmark Survival at 100 Days**

To remain informative for clinical decision-making when the median is unavailable, we report survival at a specific landmark time point. As shown in Section 3 of your code, the survival probability at 100 days is 75% ($95\% CI: 70\%-80\%$). This provides a concrete measure of patient prognosis that can be used to calibrate AI models or inform public health interventions.

---

## 4. Kaplan-Meier by Group

```{r}
#| label: km-group

km_fit <- survfit(surv_obj ~ low_ef, data = heart)

ggsurvplot(
  km_fit,
  data = heart,
  conf.int = TRUE,
  risk.table = TRUE,
  palette = "Set1",
  title = "Survival by Ejection Fraction",
  legend.labs = c("Low EF", "Normal EF")
)
```

---

## 5. Log-Rank Test

```{r}
#| label: logrank

survdiff(surv_obj ~ low_ef, data = heart)
```

The log-rank test is statistically significant (p = 0.002), indicating that survival differs between patients with low and normal ejection fraction.

---

## 6. Cox Proportional Hazards Model

```{r}
#| label: cox

cox_mod <- coxph(
  surv_obj ~ age + serum_creatinine + ejection_fraction + serum_sodium,
  data = heart,
  x = TRUE,
  y = TRUE
)

cox_mod %>%
  tbl_regression(exponentiate = TRUE) %>%
  bold_p() %>%
  add_nevent()
```

**Interpretation**:

- Age (HR = 1.05): Each additional year increases the hazard of death by 5%.
- Serum creatinine (HR = 1.39): Higher kidney dysfunction strongly increases mortality risk.
- Ejection fraction (HR = 0.96): Better cardiac function is protective.
- Serum sodium: Not statistically significant in this model.

This highlights the importance of multi-organ interactions in heart failure, consistent with a One Health perspective.

---

## 7. Diagnostics: Proportional Hazards

```{r}
#| label: ph

test_ph <- cox.zph(cox_mod)
test_ph

ggcoxzph(test_ph)
```

**Interpretation**:

None of the variables violate the proportional hazards assumption (all p-values > 0.05).

The global test is also non-significant (p = 0.165), confirming that the Cox model assumptions are satisfied.

---

## 8. Calibration (AI Perspective)

```{r}
#| label: calibration

dd <- datadist(heart)
options(datadist = "dd")

cox_rms <- cph(
  Surv(time, DEATH_EVENT) ~ age + serum_creatinine + ejection_fraction,
  data = heart,
  x = TRUE,
  y = TRUE,
  surv = TRUE
)

cal <- calibrate(
  cox_rms,
  cmethod = "KM",
  method = "boot",
  u = 100,
  m = 50,
  B = 200
)

plot(cal,
     xlab = "Predicted survival probability",
     ylab = "Observed survival probability",
     main = "Calibration Plot")

abline(0, 1, lty = 2, col = "red")
```

Calibration was performed at 100 days to align with the previously reported survival estimate.

---

## 9. Time-Dependent Covariates

```{r}
#| label: time-dependent

cox_td <- coxph(
  surv_obj ~ age + tt(age) + serum_creatinine,
  data = heart,
  tt = function(x, t, ...) x * log(t)
)

summary(cox_td)
```

**Interpretation:**

The time-dependent effect of age (tt(age)) is not statistically significant (p = 0.71), indicating that the effect of age does not vary significantly over time.

This confirms the results of the proportional hazards test and suggests that a standard Cox model is sufficient for this dataset.

---

## 10. Conclusion

This analysis demonstrates:

* Kaplan-Meier estimation
* Group comparison (log-rank test)
* Multivariate Cox modeling
* Model diagnostics
* Calibration (AI validation)
* Time-dependent effects

---

### Final Takeaway

> A good survival model is not only statistically significant,
> but also validated, calibrated, and adapted to real-world complexity.